---
title: "Analyzing complex mixed-stock fisheries with 20 stocks"
author:
- name: Sosthene Akia
  email: "sosthene.akia@dfo-mpo.ca"
  affiliation: Fisheries and Oceans Canada
- name: Alex Hanke
  email: "alex.hanke@dfo-mpo.ca"
  affiliation: Fisheries and Oceans Canada
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
vignette: "%\\\\VignetteIndexEntry{Analyzing complex mixed-stock fisheries with 20 stocks} %\\\\VignetteEngine{knitr::rmarkdown} %\\\\VignetteEncoding{UTF-8}\n"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(RHISEA)
library(tidyverse)
library(ggplot2)
```


# Introduction

This tutorial demonstrates the capability of the RHISEA package to estimate stock proportions in complex mixed-stock fisheries involving 20 distinct source populations. It guides through generating simulated classification outputs (pseudo-classes, likelihoods, phi matrix), applying the five classical HISEA estimators, and visualizing the estimates.

***

# Data Simulation of Classification Outputs

We simulate a mixture of 1000 fish from 20 stocks, generating pseudo-class assignments, likelihood matrices, and a phi matrix reflecting typical classification errors.

```{r}
set.seed(1234)

np <- 20  # number of populations
n_mixture <- 1000

# Create stock names
stocks_names <- paste0("Stock_", seq_len(np))

# Simulate pseudo-classes randomly among stocks
pseudo_classes <- sample(seq_len(np), size = n_mixture, replace = TRUE)

# Simulate likelihood matrix: concentrated probabilities near true class + noise
likelihoods <- matrix(runif(n_mixture * np, min = 0.01, max = 0.05), nrow = n_mixture, ncol = np)
for (i in seq_len(n_mixture)) {
  likelihoods[i, pseudo_classes[i]] <- likelihoods[i, pseudo_classes[i]] + 0.8
}
# Normalize rows to sum to 1
likelihoods <- likelihoods / rowSums(likelihoods)

colnames(likelihoods) <- stocks_names

# Create a phi matrix with varying diagonal and asymmetric misclassification
base_correct <- runif(np, min = 0.5, max = 0.9)
phi_matrix <- matrix(0, nrow = np, ncol = np)
for (j in 1:np) {
  phi_matrix[j, j] <- base_correct[j]
  off_diag_total <- 1 - base_correct[j]
  off_diag_indices <- setdiff(1:np, j)
  off_diag_vals <- runif(length(off_diag_indices))
  off_diag_vals <- off_diag_vals / sum(off_diag_vals) * off_diag_total
  phi_matrix[off_diag_indices, j] <- off_diag_vals
}

colnames(phi_matrix) <- stocks_names
rownames(phi_matrix) <- stocks_names

# Check columns sum to 1 (optional)
round(colSums(phi_matrix), 6)

# View phi_matrix with margins (optional)
addmargins(phi_matrix)

```

***

# Running Classic HISEA Estimators with `run_hisea_estimates()`

Using the simulated classification outputs and phi matrix, we run all five classical estimators and summarize results.

```{r}
results <- run_hisea_estimates(
  pseudo_classes = pseudo_classes,
  likelihoods = likelihoods,
  phi_matrix = phi_matrix,
  np = np,
  type = "BOOTSTRAP",
  stocks_names = stocks_names,
  export_csv = FALSE,
  verbose = TRUE
)
```

***

# Displaying and Plotting Results

We print mean estimates by estimator and stock, and visualize estimated stock proportions.

```{r plot-complex-mixed-stocks, fig.cap="Analyzing complex mixed-stock fisheries with 20 stocks"}
print(results$mean_estimates)
print(results$sd_estimates)

# Convert to tidy format for plotting
estimates_long <- as.data.frame(results$mean_estimates) %>%
  mutate(Stock = rownames(.)) %>%
  pivot_longer(-Stock, names_to = "Estimator", values_to = "Proportion")

ggplot(estimates_long, aes(x = Estimator, y = Proportion, fill = Stock)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  theme_minimal() +
  labs(title = "Estimated Stock Proportions by HISEA Estimator",
       y = "Estimated Proportion",
       x = "Estimator") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

***

# Conclusion

This tutorial illustrates that RHISEA is well suited for mixed-stock fisheries with numerous source populations. The simulation and estimation workflow scales to 20 stocks, producing robust and unbiased estimates using all classical HISEA estimators. This confirms the method’s applicability for complex, high-dimensional stock composition problems.

***
